Avoiding tipping points in the management of ecological systems: a non-parametric Bayesian approach

### 2013-10-14 inst/doc/nonparametric-bayes.tex

\begin{figure}[htbp]
\centering
\includegraphics{figure/nonparametric-bayes-statespace_plot.pdf}
\caption{Graph of the inferred Gaussian process compared to the true
process and maximum-likelihood estimated process. Graph shows the
expected value for the function $f$ under each model. Two standard
deviations from the estimated Gaussian process covariance with (light
grey) and without (darker grey) measurement error are also shown. The
training data is also shown as black points. The GP is conditioned on
(0,0), shown as a pseudo-data point.}
\end{figure}

\begin{table}[ht]
\begin{center}
\begin{tabular}{rrrr}
  \hline
 & Allen & Ricker & Myers \\ 
  \hline
DIC & 50.14 & 49.45 & 50.61 \\ 
  AIC & -26.60 & -32.07 & -29.19 \\ 
  BIC & -21.54 & -28.69 & -24.12 \\ 
   \hline
\end{tabular}
\end{center}
\end{table}

Figure 2 shows the mean inferred state space dynamics of each model
relative to the true model used to generate the data, predicting the
relationship between observed stock size (x-axis) to the stock size
after recruitment the following year. Note that in contrast to the other
models shown, the expected Gaussian process corresponds to a
distribution of curves - as indicated by the gray band - which itself
has a mean shown in black. Parameter uncertainty (not shown) spreads out
the estimates further.

While it would be straight forward to condition the GP on passing
through the origin (0,0), the estimate shown here is based only on the
observed data. The observed data from which each model is estimated is
also shown. The observations come from only a limited region of state
space corresponding to unharvested or weakly harvested system. No
observations occur at the theoretical optimum harvest rate or near the
tipping point.

@@

The resulting optimal management strategy based on each of the inferred
models is shown in Figure 2, against the optimal strategy given the true
underlying dynamics. Policies are shown in terms of target escapement,
$S_t$. Under models such as this a constant escapement policy is
expected to be optimal (Reed 1979), whereby population levels below a
certain size $S$ are unharvested, while above that size the harvest
strategy aims to return the population to $S$, resulting in the
hockey-stick shaped policies shown.

@@

The consequences of managing 100 replicate realizations of the simulated
fishery under each of the policies estimated is shown in Figure 3. As
expected from the policy curves, the structurally correct model
under-harvests, leaving the stock to vary around it's un-fished optimum.
The structurally incorrect Ricker model over-harvests the population
passed the tipping point consistently, resulting in the immediate crash
of the stock and thus derives minimal profits.

The results shown in Figures 1-3 are not unique to the simulated data or
models chosen here, but arises across a range of parameter values and
simulations as shown in the supplemental figures. The results across
this range can most easily be compared by the relative differences in
net present value realized by each of the approaches, as shown in Figure
4. The Gaussian Process most consistently realizes a value close to the
optimal solution, and importantly avoids ever driving the system across
the tipping point, which results in the near-zero value cases in the
parametric models.

@@

\subsubsection{Figure 1: Fitted Models}\label{figure-1-fitted-models}

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  All models fit the data quite well
\item
  Information criteria would pick the simple, incorrect model.
\end{itemize}

\subsubsection{Figure 2: Inferred
Policies}\label{figure-2-inferred-policies}

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Inferred policies differ substantially among models
\item
  The structurally correct model and the GP are close to the true model
\item
  alternatives are not close
\end{itemize}

\subsubsection{Figure 3: Simulated
results}\label{figure-3-simulated-results}

-

\subsubsection{Figure 4: Robustness}\label{figure-4-robustness}

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Results hold across range of parameters
\end{itemize}

(eek, distinguish better between result and discussion?)

@@

\begin{itemize}
\item
  All models are ``good fits'' to the originally observed data.
\item
  (Simple model choice immediately leads us astray)
\end{itemize}

Though simple mechanistically motivated models offer the greatest
potential to increase our basic understanding of ecological processes
(Cuddington, Fortin, and Gerber 2013; Geritz and Kisdi 2012), such
models can be not only inaccurate but misleading when relied upon in a
quantitative decision making framework.

@@

\subsubsection{1. Large uncertainty where the data is
poor}\label{large-uncertainty-where-the-data-is-poor}

@@

\subsubsection{2. Predictive accuracy where data is
good}\label{predictive-accuracy-where-data-is-good}

@@

\subsubsection{Online learning}\label{online-learning}

@@

\% latex table generated in R 3.0.1 by xtable 1.7-1 package \% Mon Sep 9
14:43:29 2013

\begin{table}[ht]
\centering
\begin{tabular}{rlrr}
  \hline
 & parameter & lower\_bound & upper\_bound \\ 
  \hline
1 & r0 & 0.00 & 10.00 \\ 
  2 & K & 0.00 & 40.00 \\ 
  3 & sigma & 0.00 & 100.00 \\ 
   \hline
\end{tabular}
\end{table}

@@

Table S2: Parameterization of the priors \% latex table generated in R
3.0.1 by xtable 1.7-1 package \% Mon Sep 9 14:43:37 2013

\begin{table}[ht]
\centering
\begin{tabular}{rlrr}
  \hline
 & parameter & lower\_bound & upper\_bound \\ 
  \hline
1 & r0 & 0.00 & 10.00 \\ 
  2 & K & 0.00 & 40.00 \\ 
  3 & theta & 0.00 & 10.00 \\ 
  4 & sigma & 0.00 & 100.00 \\ 
   \hline
\end{tabular}
\end{table}

@@

\% latex table generated in R 3.0.1 by xtable 1.7-1 package \% Mon Sep 9
14:43:45 2013

\begin{table}[ht]
\centering
\begin{tabular}{rlrr}
  \hline
 & parameter & lower\_bound & upper\_bound \\ 
  \hline
1 & r0 & 0.00 & 10.00 \\ 
  2 & K & 0.00 & 40.00 \\ 
  3 & theta & 0.00 & 10.00 \\ 
  4 & sigma & 0.00 & 100.00 \\ 
   \hline
\end{tabular}
\end{table}

@@

Cuddington, Kim M., M. J. Fortin, and L. R. Gerber. 2013.
``Process-based Models Are Required to Manage Ecological Systems in a
Changing World.'' \emph{Ecosphere} 4 (February): 1--12.
doi:10.1890/ES12-00178.1.
\url{http://www.esajournals.org/doi/abs/10.1890/ES12-00178.1}.

@@

Geritz, Stefan A. H., and Eva Kisdi. 2012. ``Mathematical Ecology: Why
Mechanistic Models?'' \emph{Journal of Mathematical Biology} 65 (6-7)
(dec): 1411--5. doi:10.1007/s00285-011-0496-3.
\url{http://www.ncbi.nlm.nih.gov/pubmed/22159789}.

>>>>

All models fit the observed data rather closely and with relatively
small uncertainty, as illustrated in the posterior predictive curves in
Figure 1. Figure 1 shows the training data of stock sizes observed over
time as points, overlaid with the step-ahead predictions of each
estimated model using the parameters sampled from their posterior
distributions. Each model manages to fit the observed data rather
closely. Compared to the expected value of the true model most estimates
appear to overfit, predicting fluctuations that are actually due purely
to stochasticity in growth rate. Model-choice criteria shown in Table 1
penalize more complex models and show a slight preference for the
simpler Ricker model over the more complicated alternate stable state
models (Allen and Myers). Details on MCMC estimates for each model,
traces, and posterior distributions can be found in the appendix.

\begin{table}[ht]
\begin{center}
\begin{tabular}{rrrr}
  \hline
 & Allen & Ricker & Myers \\ 
  \hline
DIC & 50.14 & 49.45 & 50.61 \\ 
  AIC & -24.60 & -30.07 & -27.19 \\ 
  BIC & -17.85 & -25.00 & -20.44 \\ 
   \hline
\end{tabular}
\end{center}
\end{table}

\begin{figure}[htbp]
\centering
\includegraphics{figure/nonparametric-bayes-statespace_posteriors.pdf}
\caption{Graph of the inferred Gaussian process compared to the true
process and maximum-likelihood estimated process. Graph shows the
expected value for the function $f$ under each model. Two standard
deviations from the estimated Gaussian process covariance with (light
grey) and without (darker grey) measurement error are also shown. The
training data is also shown as black points. The GP is conditioned on
(0,0), shown as a pseudo-data point.}
\end{figure}

The mean inferred state space dynamics of each model relative to the
true model used to generate the data is shown in Figure 2, predicting
the relationship between observed stock size (x-axis) to the stock size
after recruitment the following year. Note that in contrast to the other
models shown, the expected Gaussian process corresponds to a
distribution of curves - as indicated by the gray band - which itself
has a mean shown in black. Parameter uncertainty (not shown) spreads out
the estimates further.\\The observed data from which each model is
estimated is also shown. The observations come from only a limited
region of state space corresponding to unharvested or weakly harvested
system. No observations occur at the theoretical optimum harvest rate or
near the tipping point.

\begin{figure}[htbp]
\centering
\includegraphics{figure/nonparametric-bayes-out_of_sample_predictions.pdf}
\caption{plot of chunk out\_of\_sample\_predictions}
\end{figure}

@@

Despite the similarities in model fits to the observed data, the
policies inferred under each model differ widely, as shown in Figure
3.\\Policies are shown in terms of target escapement, $S_t$. Under
models such as this a constant escapement policy is expected to be
optimal (Reed 1979), whereby population levels below a certain size $S$
are unharvested, while above that size the harvest strategy aims to
return the population to $S$, resulting in the hockey-stick shaped
policies shown. Only the structurally correct model (Allen model) and
the GP produce policies close to the true optimum policy (where both the
underlying model structure and parameter values are known without
error).

@@

The consequences of managing 100 replicate realizations of the simulated
fishery under each of the policies estimated is shown in Figure 4. As
expected from the policy curves, the structurally correct model
under-harvests, leaving the stock to vary around it's un-fished optimum.
The structurally incorrect Ricker model over-harvests the population
passed the tipping point consistently, resulting in the immediate crash
of the stock and thus derives minimal profits.

These results are robust across a range of stochastic realizations,
models, and parameter values.\\The results across this range can most
easily be compared by using the relative differences in net present
value realized by each of the model, as shown in Figure 5. The BNP-SDP
approach most consistently realizes a value close to the optimal
solution, and importantly avoids ever driving the system across the
tipping point, which results in the near-zero value cases in the
parametric models.

@@

@@

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  All models are ``good fits'' to the originally observed data.
\item
  (Simple model choice immediately leads us astray)
\end{itemize}

@@

\subsubsection{Traditional model-choice approaches can be positively
misleading.}\label{traditional-model-choice-approaches-can-be-positively-misleading.}

These results illustrate that model-choice approaches would be
positively misleading -- supporting simpler models that cannot express
tipping point dynamics merely on account of them being similar. As the
data shown comes only from the basin of attraction near the unfished
equilibrium, near which all of the models are approximately linear and
approximately identical.

Model choice approaches trade off model complexity and fit to the data.
When the data come from a limited region of state-space -- as is
necessarily the case whenever there is a potential concern about tipping
point dynamics -- simpler models can fit just as well and will tend to
outperform more complex ones. This approach would be appropriate when
the dynamics can be expected to remain in the region of the training
data; for instance, if we only considered the forecasting accuracy of
the unfished population dynamics under each model.

In contrast, the decision-maker's problem of setting appropriate harvest
levels cannot exclude regions of state-space outside the observed range
when integrating over all possible decisions to find the optimal choice.
Such problems are not constrained to fisheries management but ubiquitous
across ecological decision-making and conservation where the greatest
concerns involve entering previously unobserved regions of state-space
-- whether that is the collapse of a fishery, the spread of an invasive,
or the loss of habitat.

\subsubsection{BNP-SDP expresses larger uncertainty in regions where the
data are
poor}\label{bnp-sdp-expresses-larger-uncertainty-in-regions-where-the-data-are-poor}

@@

\subsubsection{BNP-SDP has good predictive accuracy where data are
good}\label{bnp-sdp-has-good-predictive-accuracy-where-data-are-good}

@@

\subsubsection{Real-time learning}\label{real-time-learning}

@@

\begin{table}[ht]
\begin{center}
\begin{tabular}{rlrr}
  \hline
 & parameter & lower\_bound & upper\_bound \\ 
  \hline
1 & r0 & 0.00 & 10.00 \\ 
  2 & K & 0.00 & 40.00 \\ 
  3 & sigma & 0.00 & 100.00 \\ 
   \hline
\end{tabular}
\end{center}
\end{table}

@@

Table S2: Parameterization of the priors

\begin{table}[ht]
\begin{center}
\begin{tabular}{rlrr}
  \hline
 & parameter & lower\_bound & upper\_bound \\ 
  \hline
1 & r0 & 0.00 & 10.00 \\ 
  2 & K & 0.00 & 40.00 \\ 
  3 & theta & 0.00 & 10.00 \\ 
  4 & sigma & 0.00 & 100.00 \\ 
   \hline
\end{tabular}
\end{center}
\end{table}

@@

\begin{table}[ht]
\begin{center}
\begin{tabular}{rlrr}
  \hline
 & parameter & lower\_bound & upper\_bound \\ 
  \hline
1 & r0 & 0.00 & 10.00 \\ 
  2 & K & 0.00 & 40.00 \\ 
  3 & theta & 0.00 & 10.00 \\ 
  4 & sigma & 0.00 & 100.00 \\ 
   \hline
\end{tabular}
\end{center}
\end{table}

@@

@@
